Batch Effects in the 1000 Genomes Project Data Confounds Modern Genomics Studies

### 2018-09-13 legacy.tex
finalized stats methods with Rick, and added reference

The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}). We wanted a test statistic to represent the association across all populations, to do that we summed across the deviations from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.

@@

\subsection{Genome wide simple logistic regression analysis}
We fitted a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}. The simple logistic model can be expressed as such.

\begin{align*}
{logit}{E(Y_{s'p'i})} &= \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{p'i}
\\
\textit{for}\quad i &= 1,\hdots, n_{p'}
\end{align*}
where $Y_{spi}$ is a Bernouli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$. We run a total of $K (= \sum_{s=1}^S P_s)$ regressions where $ P_s$, ranging from 2 to 26, is the number of populations with a segregating variant for genomic locus $s$.

For each combination of of the 26 populations and for each locus, we formulated a null hypothesis that the quality variable is not associated with genotype, that is, the slope  coefficient of the simple logistic regression is zero. $H_{0}: \beta_{1}=0$.
To test the null hypothesis, we used the chi square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of the average quality of mapped bases. 
The deviance test statistic is approximately chi-square distributed with one degrees of freedom.  
The deviations from the null model for each test was used to compute the chi-squared \textit{p}-value for each SNP for each population.
To test whether the quality variable is associated with genotype across all populations, we used the sum of the deviances across the populations.
However, since not all SNPs are present in all populations, when computing the \textit{p}-values for the summed deviations, the degrees of freedom were adjusted to the number of populations tested.

@@

\caption{Manhattan plot of the -log10 p values for the combined test using the deviations from each population of the 1kGP. 
There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{Manhattan}
\end{figure}

>>>>

The GWAS approach similarly identified mapping-quality--associated SNPs in 24 of the 26 populations in the 1kGP.
Over 3517 variants were independently associated to low quality in at least two populations  (Figure \ref{OverLap}). We wanted a test statistic to represent the association across all populations, to do that we summed across the deviances from each population (adjusting the degrees of freedom for the chi-squared test according to the number of populations with a segregating variant) and identified 3990 statistically significant SNPs, as shown on Figure \ref{Manhattan}.

@@

\subsection{Genome wide simple logistic regression analysis}
We ran a simple logistic regression independently for each combination of genomic locus $s$ and population $p$ using the R glm package\citep{RDevelopmentCoreTeam2016}. The model we fitted to each subset of data can be expressed as such.
\begin{align*}
{logit}{E(Y_{s'p'i})} &= \beta_{0}^{(sp)} + \beta_{1}^{(sp)} Q_{p'i}
\\
\textit{for}\quad i &= 1,\hdots, n_{p'}
\end{align*}
Here $Y_{spi}$ is a Bernouli random variable denoting the genotype of individual $i$ in population $p$ at genomic locus $s$. 
In the model, the regression coefficient $\beta_{0}^{(sp)}$ represents the baseline allele frequency in population $p$, the slope coefficient $\beta_{1}^{(sp)}$ denotes the association to average quality of mapped bases $Q_{p}$ to genotype $Y_{sp}$. 
We run a total of $K (= \sum_{s=1}^S P_s)$ regressions, where $S$ is the \todo{total number of genomic loci} , and $P_s$ is the number of populations with a segregating variant for locus $s$. In the data, $P_s$ ranges from 2 to 26 since not all SNPs were present in all populations and loci where only a single variant was observed were filtered out.

For each of the $K$ regressions, we formulated a null hypothesis $H_{0}: \beta_{1}^{(sp)}=0$. That is, the average quality of mapped bases $Q_{p}$ variable is not associated with genotype $Y_{sp}$.
In other words, the slope coefficient of the simple logistic regression is zero.
To test the null hypothesis, we used the chi-square likelihood ratio test statistic (the deviance) as a measure of the marginal importance of adding $Q_{p}$ in the model. 
The deviance test statistic is approximately chi-square distributed with one degrees of freedom.

To obtain a population-wide measure of association for each locus across its respective $P_s$ populations, we summed the $P_s$ deviances for each $s$. The resulting test statistic $T_s$ is chi-squared distributed with $P_s$ degrees of freedom (i.e. $T_s \sim \chi^2_{P_s}$), see Figure \ref{Deviances}.
The observed summed deviances were used to compute the \textit{p}-value for each locus.

Given the large number of \textit{p}-values of which there are $S$ of them, the large proportion of expected null hypotheses and the positive dependencies across the genome, we used the two-stage Benjamini \& Hochberg step-up FDR-controlling procedure to adjust the \textit{p}-values
\citep{Benjamini2006}.
By using a very conservative nominal Type-I error rate $\alpha = 0.01$, a total of 3990 loci were found to be statistically significance. See supplementary file {here}. \luke{make this file}

@@

\caption{Manhattan plot of the -log10 p values for the combined test using the deviances from each population of the 1kGP. 
There are 3990 variants that reach p values greater than $ p < 0.01$ after performing a two-stage Benjamini and Hochberg FDR adjustment. 
The circles ( o ) are SNPs that reached values greater than 20, for clarity we implemented hard ceiling at 20.}
  \label{Manhattan}
\end{figure}
